---
title: "Distribution shift metrics"
format: html
editor: visual
execute:
  echo: true
  eval: true
---

# Goals:

-   Gain exposure to calculating a variety of distribution shift metrics with sdmTMB (center of gravity, effective area occupied, range edges, covariate weighted averages)
-   Explore how spatially varying coefficients can reveal spatially structured trends
-   Calculate subregional population indices

# Setup

Load packages and set plotting defaults so figures are consistent throughout.

```{r}
#| echo: false
#| message: false
#| warning: false
library(sdmTMB)
library(dplyr)
library(ggplot2)
options(ggplot2.continuous.colour = "viridis")
options(ggplot2.continuous.fill = "viridis")
options(ggplot2.discrete.colour = RColorBrewer::brewer.pal(8, "Set2"))
options(ggplot2.discrete.fill = RColorBrewer::brewer.pal(8, "Set2"))
theme_set(theme_light())
```

In this vignette, we'll demonstrate calculating distribution shift metrics for Pacific Spiny Dogfish (*Squalus suckleyi*) on the west coast of Vancouver Island.

# Dataset

We'll use the built-in `dogfish` dataset, which contains data from DFO's Synoptic West Coast Vancouver Island bottom trawl survey. The dataset includes catch weights, depth, and area swept. It is included with sdmTMB.

The dataset includes projected spatial coordinates (`X`, `Y`) (UTM zone 9 KM), which we will use to define the spatial axis for distribution metrics. The prediction grid includes the same covariates (e.g., depth) so we can make model-based predictions across the domain.

```{r glimpse-dogfish}
glimpse(dogfish)
```

For prediction, we'll use the `wcvi_grid`, which provides a spatial grid covering the survey area. This is also in the sdmTMB package.

```{r glimpse-grid}
glimpse(wcvi_grid)
```

# Fitting a spatiotemporal model

First, we'll construct a mesh for the spatial random effects:

```{r mesh, fig.asp=0.8}
mesh <- make_mesh(dogfish, c("X", "Y"), cutoff = 10)
```

Next, we'll fit a spatiotemporal model for dogfish density using a delta generalized-gamma family (Dunic et al. 2025). This models the positive catches with a generalized gamma distribution that works well for Pacific Dogfish given their occasional giant outlying catch values. A simpler Tweedie or delta-gamma family could have been used too. We'll include depth as a predictor using a quadratic effect:

```{r fitmodel}
#| results: hide
fit <- sdmTMB(
  catch_weight ~ poly(log(depth), 2),
  data = dogfish,
  mesh = mesh,
  family = delta_gengamma(type = "poisson-link"),
  spatial = "on",
  time = "year",
  spatiotemporal = "IID",
  silent = FALSE
)
```

```{r}
sanity(fit)
fit
```

The model includes spatial and spatiotemporal random fields, allowing density to vary across space and time. The depth effect captures the strong association between dogfish density and bathymetry.

# Making predictions

To calculate range edges, we need to make predictions on a spatial grid that covers the area of interest. We'll replicate the `wcvi_grid` for each year in the dataset:

```{r predict}
years <- sort(unique(dogfish$year))
nd <- replicate_df(wcvi_grid, "year", years)
```

We will also keep the TMB object so downstream functions (`get_cog()` etc.) have the TMB object to work with.

```{r}
predictions <- predict(fit, newdata = nd, return_tmb_object = TRUE)
```

# Center of gravity

The center of gravity (COG) gives the density-weighted mean location of the distribution. We can summarize it as a single point per year, with uncertainty intervals for both the x and y coordinates. `format = "wide"` makes it easier to plot here with ggplot.

Although this idea has been around since at least the 1980s, applying it to predictions from a GMRF model was first described in Thorson et al. (2016). See Woillez et al. (2007) to start diving into older references.

```{r}
cog <- get_cog(predictions, format = "wide")
cog
```

```{r}
ggplot(cog, aes(est_x, est_y, colour = year)) +
  geom_point() +
  geom_linerange(aes(xmin = lwr_x, xmax = upr_x)) +
  geom_linerange(aes(ymin = lwr_y, ymax = upr_y))

ggplot(cog, aes(year, est_x)) +
  geom_point() +
  geom_linerange(aes(ymin = lwr_x, ymax = upr_x))

ggplot(cog, aes(year, est_y)) +
  geom_point() +
  geom_linerange(aes(ymin = lwr_y, ymax = upr_y))
```

# Density-weighted habitat covariate

This metric summarizes how the distribution shifts with respect to a habitat covariate. The concept si the same as the center of gravity but applied to variables other than latitude and longitude. 

Here, we'll use depth. We flip the sign in the plot so deeper values appear lower on the y-axis.

```{r}
wa <- get_weighted_average(predictions, vector = nd$depth)
wa
ggplot(wa, aes(year, -est)) + geom_line() +
  geom_ribbon(aes(ymin = -lwr, ymax = -upr), alpha = 0.4)
```

We could also do this with temperature and potentially compare to the density-weighted mean temperature with an average spatial distribution.

See Davidson et al. (2025) for an expanded version of this example and Ward et al. (2024) for a related example with niche width.

# Effective area occupied

The effective area occupied (EAO) is a distributional metric related to concentration. It can decrease when a population becomes more spatially clustered, and increase when the population spreads out. See Thorson et al. (2016) for an explanation of the metric as calculated here. See `exercises/understanding-eao.R` for a walk through.

It's saying: if the population were spread uniformly at the expected density experienced by a randomly chosen individual then it would need an area of XX km^2^ to contain the total biomass that year.

```{r}
eao <- get_eao(predictions, area = 4)
eao
ggplot(eao, aes(year, est)) + geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.4) +
  ylim(0, NA) +
  ylab("Effective area occupied (km^2)")
```

# Range edges

The `get_range_edge()` function calculates range edges as density-weighted quantiles along a spatial axis (e.g., latitude, longitude, or depth). This approach follows methods used in Fredston et al. (2021) from VAST.

Range edges are calculated by:

1.  Ordering spatial locations along a user-specified axis
2.  Calculating the cumulative proportion of total density along that axis
3.  Finding positions where the cumulative proportion equals target quantiles
4.  Using simulation from the joint precision matrix to quantify uncertainty

The `predict()` function with `nsim > 0` returns a matrix where each column represents one simulation draw from the joint precision matrix.

```{r}
set.seed(123)
pred_sims <- predict(fit, newdata = nd, nsim = 300)
```

Now we can calculate range edges along the latitude axis (Y coordinate). By default, `get_range_edge()` calculates the 2.5% and 97.5% quantiles, representing the lower and upper range edges. Here, we'll also add in the median (0.5) to find the center of the distribution.

```{r range-edges}
edges <- get_range_edge(pred_sims, axis = nd$Y, quantiles = c(0.05, 0.50, 0.95))
head(edges)
```

The output includes:

-   `year`: the time slice
-   `quantile`: the quantile value (0.05 for lower edge, 0.50 for the median, 0.95 for upper edge)
-   `est`: the estimated position along the axis (latitude in this case)
-   `lwr` and `upr`: 90% confidence intervals for each quantile
-   `se`: standard error for each quantile

## Visualizing range edges

We can plot how the range edges change over time:

```{r plot-custom-edges, fig.width=7, fig.asp=0.5}
ggplot(edges, aes(year, est)) +
  geom_line(aes(colour = factor(quantile))) +
  geom_ribbon(
    aes(ymin = lwr, ymax = upr, fill = factor(quantile)),
    alpha = 0.2
  ) +
  labs(
    x = "Year",
    y = "Latitude (UTM km)",
    colour = "Quantile",
    fill = "Quantile"
  )
```

This plot shows how the northern and southern range edges of dogfish have shifted over time, with uncertainty bands reflecting sampling and estimation uncertainty. The median line represents the center of the distribution.

## Custom analyses with simulation draws

For custom analyses, you can access the raw simulation draws:

```{r sims}
edges_sims <- get_range_edge(pred_sims, axis = nd$Y, return_sims = TRUE, quantiles = c(0.95))
head(edges_sims)
```

This returns all simulation draws in long format, which can be useful for:

-   Custom uncertainty quantification
-   Calculating probabilities of range shifts
-   Comparing range edges between models or scenarios

For example, we could calculate the probability that the upper range edge (northern edge) has shifted northward between two time periods:

```{r prob-shift}
# Extract simulations for upper edge in first and last year
upper_first <- edges_sims |>
  filter(year == min(year))

upper_last <- edges_sims |>
  filter(year == max(year))

# Calculate shift for each simulation
shifts <- upper_last$.value - upper_first$.value[match(upper_last$.iteration, upper_first$.iteration)]

# Probability of northward shift
prob_north <- mean(shifts > 0)
cat("Probability of northward shift:", round(prob_north * 100, 1), "%\n")
```

We can plot that distribution:

```{r prob-shift-hist}
ggplot(data.frame(shifts = shifts), aes(shifts)) + geom_histogram()
```

If anything, the northern range has contracted here, which is consistent with previous research on this stock (Ward et al. 2024).

## Other axes

While we've demonstrated using latitude (Y), range edges can be calculated along any spatial axis. Choices include:

-   **Longitude (X)**: for east-west range shifts
-   **Along shoreline**: for range edges along a shoreline that are not well represented by latitude or longitude alone
-   **Coastal distance**: for offshore/onshore distribution patterns
-   **Depth**: for depth range edges (e.g., shallow vs. deep distribution limits)
-   **Temperature**: if you have environmental covariates in your prediction grid

Simply provide the appropriate vector to the `axis` argument.

E.g. here we can calculate the metrics for depth:

```{r}
depth_edges <- get_range_edge(pred_sims, axis = nd$depth, quantiles = c(0.05, 0.5, 0.95))

ggplot(depth_edges, aes(year, -est, colour = as.factor(quantile))) +
  geom_line() +
  geom_ribbon(
    aes(ymin = -lwr, ymax = -upr, fill = as.factor(quantile)),
    alpha = 0.2,
    colour = NA
  ) +
  labs(
    x = "Year",
    y = "Depth (m)",
    colour = "Quantile",
    fill = "Quantile"
  ) +
  scale_colour_discrete(labels = c("5%", "50%", "95%")) +
  scale_fill_discrete(labels = c("5%", "50%", "95%"))
```

So, in particular, the dogfish are moving deeper at the deepest extent of their distribution.

# Spatially varying trends

Spatially varying coefficients (SVCs) let trends differ across space. Here we include a linear decade effect and allow its coefficient to vary spatially, highlighting where the strongest increases or decreases occur.

See Barnett et al. 2021 and Davidson et al. 2025 for examples.

```{r}
#| results: hide
dog <- dogfish
dog$decade <- (dog$year - mean(dog$year))/10 # center and convert to decades
fit_svc <- sdmTMB(
  catch_weight ~ poly(log(depth), 2) + decade,
  spatial_varying = ~ 0 + decade,
  data = dog,
  mesh = mesh,
  family = delta_gengamma(type = "poisson-link"),
  spatial = "on",
  time = "year",
  spatiotemporal = "off", # or "iid"
  silent = FALSE
)
```

```{r}
sanity(fit_svc)
```

```{r}
fit_svc
```

```{r}
# we only need one year for visualization here
nd_one_year <- filter(nd, year == max(nd$year))

# make sure to center year by the same mean as in fitting!
nd_one_year$decade <- (nd_one_year$year - mean(dog$year)) / 10

pred_svc <- predict(fit_svc, newdata = nd_one_year)
```

To get the overall spatial trends, we need to combine the fixed and random effects from both linear predictors of the delta model. Because we used the Poisson-link, both are in log space and we can simply add them. 

We **could not** do this with a standard delta model that has a logit and log link. We need to use the Poisson-link delta *or* a model with one linear predictor (e.g., Tweedie, NB2).

```{r}
fe1 <- coef(fit_svc, model = 1)[["decade"]] # fixed effect from 1st linear predictor
fe2 <- coef(fit_svc, model = 2)[["decade"]] # fixed effect from 2nd linear predictor
# we could have also used tidy() ?tidy.sdmTMB

# the overall SVC is the fixed effects plus the random effects
pred_svc <- mutate(pred_svc, svc = fe1 + fe2 + zeta_s_decade1 + zeta_s_decade2)
```

```{r}
ggplot(pred_svc, aes(X, Y, fill = exp(svc))) + geom_tile() +
  scale_fill_gradient2(trans = "log10") +
  labs(fill = "Multiplicative\nchange\nper decade")
```

# Subregional population indices

We can calculate population indices for portions of a survey grid by filtering our survey grey, predicting on it, and then integrating the biomass (or abundance) over that filtered grid. This lets us ask questions about how biomass or abundance has been changing over smaller spatial regions.

We could also do this for center of gravity as in Barnett et al. (2021).

```{r}
# visualize our grid
mutate(nd, north_half = Y > 5450) |> 
  ggplot(aes(X, Y, fill = north_half)) + geom_tile()

# split our grid
grid_north <- filter(nd, Y > 5450)
grid_south <- filter(nd, Y <= 5450)

# predict on each half
pred_north <- predict(fit, newdata = grid_north, return_tmb_object = TRUE)
pred_south <- predict(fit, newdata = grid_south, return_tmb_object = TRUE)

# integrate density over the grid
index_north <- get_index(pred_north, area = 4, bias_correct = TRUE)
index_south <- get_index(pred_south, area = 4, bias_correct = TRUE)

# visualize the index
bind_rows(
  mutate(index_north, subregion = "North WCVI"),
  mutate(index_south, subregion = "South WCVI")
) |> 
  ggplot(aes(year, y = est, ymin = lwr, ymax = upr)) + 
  geom_ribbon(aes(fill = subregion), alpha = 0.2) +
  geom_line(aes(colour = subregion))
  # scale_y_log10()
```

# References

Barnett, L.A.K., Ward, E.J., and Anderson, S.C. 2021. Improving estimates of species distribution change by incorporating local trends. Ecography 44(3): 427–439. <https://doi.org/10.1111/ecog.05176>

Davidson, L.N.K., English, P.A., King, J., Grant, P., Taylor, I.G., Barnett, L., Gertseva, V., Tribuzio, C.A., and Anderson, S.C. 2025. Mystery of the disappearing dogfish: transboundary analyses reveal steep population declines across the Northeast Pacific with little evidence for regional redistribution. Fish and Fisheries. <https://doi.org/10.1111/faf.70028>.

Dunic, J.C., Conner, J., Anderson, S.C., and Thorson, J.T. 2025. The generalized gamma is a flexible distribution that outperforms alternatives when modelling catch rate data. ICES Journal of Marine Science 82(4): fsaf040. <https://doi.org/10.1093/icesjms/fsaf040>

Fredston, A. L., Pinsky, M., Selden, R. L., Szuwalski, C., Thorson, J. T., Gaines, S. D., & Halpern, B. S. (2021). Range edges of North American marine species are tracking temperature over decades. *Global Change Biology*, 27(13), 3145-3156. <https://doi.org/10.1111/gcb.15614>

Thorson, J.T., Rindorf, A., Gao, J., Hanselman, D.H., and Winker, H. 2016. Density-dependent changes in effective area occupied for sea-bottom-associated marine fishes. Proceedings of the Royal Society B: Biological Sciences 283(1840): 20161853. Royal Society. <https://doi.org/10.1098/rspb.2016.1853>.

Thorson, J.T., Pinsky, M.L., and Ward, E.J. 2016. Model-based inference for estimating shifts in species distribution, area occupied and centre of gravity. Methods Ecol Evol 7(8): 990–1002. <https://doi.org/10.1111/2041-210X.12567>.

Ward, E.J., Anderson, S.C., Barnett, L.A.K., English, P.A., Berger, H.M., Commander, C.J.C., Essington, T.E., Harvey, C.J., Hunsicker, M.E., Jacox, M.G., Johnson, K.F., Large, S., Liu, O.R., Richerson, K.E., Samhouri, J.F., Siedlecki, S.A., Shelton, A.O., Somers, K.A., and Watson, J.T. 2024. Win, lose, or draw: Evaluating dynamic thermal niches of northeast Pacific groundfish. PLOS Climate 3(11): e0000454. Public Library of Science. <https://doi.org/10.1371/journal.pclm.0000454>

Woillez, M., Poulard, J.-C., Rivoirard, J., Petitgas, P., and Bez, N. 2007. Indices for capturing spatial patterns and their evolution in time, with application to European hake (*Merluccius merluccius*) in the Bay of Biscay. ICES J Mar Sci 64(3): 537–550. <https://doi.org/10.1093/icesjms/fsm025>.
